Abstractions for Type-Level Programming

### 2022-04-23 article.tex
More diff

@@

\subsection{Comparison to Generalized Singleton Types}

In \Cref{chap:generalizing-singleton-types}, we presented a generalization of singleton types for type-level programing.
Generalized singletons and match types share some similarities, such as the notion of disjointness which plays an important role in both features.
The two features relate on deeper level: we can explain match types in terms of generalized singletons.
For example, the we can give an alternative definition of |Elem| match type using a generalized singleton:
%
\begin{lstlisting}
type Elem[X] = { (_: X) match
  case _: String => (_: Char)
  case _: List[t] => (_: Elem[t])
  case _: Any => (_: X)
}
\end{lstlisting}
%
This definition makes the term/type correspondence evident: the |Elem[X]| type reduces in the way than the corresponding match expression.

@@

Scala tuples, for example, fall into this category.
|Tuple2|, the class for pairs, is defined as follows:
%
\begin{lstlisting}
case class Tuple2[+T1, +T2](_1: T1, _2: T2)
\end{lstlisting}
%
Given two instantiations of this this type, |Tuple2[Int,X]| and |Tuple2[String,X]|, although |Tuple2[Nothing,X]| is a subtype of both, there is no runtime value of type |Tuple2[Nothing,X]| (since |Nothing| is uninhabited), and, as a result, those two types are disjoint.
Our disjointness algorithm implements this kind of reasoning to conclude disjointness in the presence of covariant type parameters.
\begin{diff}
There is no such reasoning available for contravariant types: our algorithm can never conclude that |G[X]| and |G[Y]| are disjoint for a contravariant type |G|.
\end{diff}

@@

\begin{diff}
For example, consider the following function definition:
%
\begin{lstlisting}
def foo[X](x: X) = x match
  case x: Int => x.toString
  case x: String => x.toInt
\end{lstlisting}
%
In the absence of a result type annotation, Scala's type inference will first assign \lstinline!String|Int! for the match expression, and then immediately widen that union to the least upper bound of left- and right-hand-side, |Any|, which will then be used as |foo|'s (inferred) result type.
To obtain a precise result type for this method, users should annotate it explicitly:
%
\begin{lstlisting}
type Foo[X] = X match
  case Int => String
  case String => Int

@@

We implement reductions along a given list of indices with logic similar to two nested loops.
The outer loop, |Loop|, traverses the shape and counts the current index.
The inner loops are implemented using the standard |Contains| and |Remove| operations.
When |Loop| reaches the end of the list, if there are still axes to remove, these are out of bounds for the initial shape: the match type intentionally gets stuck in such cases.
%
\reduceType
\reduceLoop

>>>>

\subsection{Comparison to Generalized Singleton Types}
\begin{diff}
In \Cref{chap:generalizing-singleton-types}, we presented a generalization of singleton types for type-level programing.
Generalized singletons and match types share some similarities, such as the notion of disjointness, which plays an important role in both features.
The two features relate on a deeper level: we can explain match types in terms of generalized singletons.
For example, we can give an alternative definition of the |Elem| match type using a generalized singleton:
%
\begin{lstlisting}
type Elem[X] = { (_: X) match
  case _: String => (_: Char)
  case _: List[t] => (_: Elem[t])
  case _: Any => (_: X) }
\end{lstlisting}
%
This definition makes the term/type correspondence evident: the |Elem[X]| type reduces analogously to the corresponding match expression.
\end{diff}

@@

@@

Scala tuples, for example, fall into this category.
|Tuple2|, the class for pairs, is defined as follows: |case class Tuple2[+T1, +T2](_1: T1, _2: T2)|.
\begin{diff}
Given two instantiations of this this type, |Tuple2[Int,X]| and |Tuple2[String,X]|, although |Tuple2[Nothing,X]| is a subtype of both, there is no runtime value of type |Tuple2[Nothing,X]| (since |Nothing| is uninhabited), and, as a result, those two types are disjoint.
Our disjointness algorithm implements this kind of reasoning to conclude disjointness in the presence of covariant type parameters.
There is no such reasoning available for contravariant types: our algorithm can never conclude that |G[X]| and |G[Y]| are disjoint for a contravariant type |G|.
\end{diff}

@@

\begin{diff}
For example, consider the following function definition:

\begin{minipage}{\linewidth}
\begin{lstlisting}
def foo[X](x: X) = x match
  case x: Int => x.toString
  case x: String => x.toInt
\end{lstlisting}
\end{minipage}
%
In the absence of a result type annotation, Scala's type inference will first assign \lstinline!String|Int! for the match expression, and then immediately widen that union to the least upper bound of left- and right-hand-side, |Any|, which will then be used as |foo|'s (inferred) result type.
To obtain a precise result type for this method, users should annotate it explicitly:
%
\begin{lstlisting}
type Foo[X] = X match
  case Int => String
  case String => Int

@@

We implement reductions along a given list of indices with logic similar to two nested loops.
The outer loop, |Loop|, traverses the shape and counts the current index.
The inner loops are implemented using the standard |Contains| and |Remove| operations.
When |Loop| reaches the end of the list, if there are still axes to remove, these are out of bounds for the initial shape: the match type intentionally gets stuck in such cases.

\begin{minipage}{\linewidth}
\reduceType
\reduceLoop
\end{minipage}

### 2022-04-23 symposium.tex
More diff

Capturing groups allow programmers to extract substrings matched by parts of a regular expression.
For example, the regex |"A(B)?"| matches both |"A"| and |"AB"|.
In the first case, the capturing group is empty, and in the second case, the capturing group contains |"B"|.

The Scala standard library contains a package to manipulate regular expressions.
This package is based on Java's implementation, and thus benefits from the high performance of the JVM's regex engine.
The Scala implementation innovates in its presentation: it improves upon Java's solution by providing an ergonomic API based on pattern matching that is both elegant and concise.
The package's documentation starts with the following example:

@@

This \paper is structured as follows.
In \Cref{sec:background}, we give an introduction to match types and generic tuples, two recent additions to Scala's type system which we rely on in our implementation.
In \Cref{sec:architecture},~\ref{sec:type-level} and~\ref{sec:term-level}, we present our library for type-safe regular expressions.
In \Cref{sec:evaluation}, we evaluate the performance of our implementation in terms of compilation times and execution times.
In \Cref{sec:related}, we discuss related work.
We conclude the \paper in \Cref{sec:conclusion}.

@@

\section{Background}
\label{sec:background}

@@

\tupleReverseA
\vspace{-3pt}
\tupleReverseB

@@

We present our library for type-safe regular expressions in terms of 3 components:

@@

Stephanie Weirich presented on several occasions a regular expression library as an example of a dependently-typed program written in Haskell \citep{weirich2014examples}.
The library provides safety guarantees similar to ours, but from the slightly different angle of named-capturing groups.
An important conceptual difference between both approaches is ours only focuses on type checking (we rely on Java's regular expressions at run-time), whereas hers also includes a fully-fledged regular expression engine.

@@

The safe-regex~\citep{leonhard2021safe} Rust library takes a different approach towards the same goal: it uses Rust macros to statically compile regular expressions and generate arity-specific extractors.
Performance-wise, macros improve over the status quo since they remove the need to compile regular expressions at runtime.
Safety-wise, code generation provides a straightforward solution to safe capturing group extraction.

>>>>

\begin{diff}
Capturing groups allow programmers to extract substrings matched by parts of a regular expression.
For example, the regex |"A(B)?"| matches both |"A"| and |"AB"|.
In the first case, the capturing group is empty, and in the second case, the capturing group contains |"B"|.
\end{diff}

The Scala standard library contains a package to manipulate regular expressions.
This package is based on Java's implementation, and thus benefits from the high performance of the JVM's regex engine.
\begin{diff}
The Scala implementation innovates in its presentation: it improves upon Java's solution by providing an ergonomic API based on pattern matching that is both elegant and concise.
The package's documentation starts with the following example:
\end{diff}

@@

\begin{diff}
This \paper is structured as follows.
In \Cref{sec:background}, we give an introduction to match types and generic tuples, two recent additions to Scala's type system which we rely on in our implementation.
In \Cref{sec:architecture},~\ref{sec:type-level} and~\ref{sec:term-level}, we present our library for type-safe regular expressions.
In \Cref{sec:evaluation}, we evaluate the performance of our implementation in terms of compilation times and execution times.
In \Cref{sec:related}, we discuss related work.
We conclude the \paper in \Cref{sec:conclusion}.
\end{diff}

@@

\begin{diff}
\section{Background}
\label{sec:background}

@@

\tupleReverseA
\vspace{-3pt}
\tupleReverseB
\end{diff}

@@

\begin{diff}
We present our library for type-safe regular expressions in terms of 3 components:
\end{diff}

@@

Stephanie Weirich presented on several occasions a regular expression library as an example of a dependently-typed program written in Haskell \citep{weirich2014examples}.
The library provides safety guarantees similar to ours, but from the slightly different angle of named-capturing groups.
\begin{diff}
An important conceptual difference between the two approaches is that ours only focuses on type checking (we rely on Java's regular expressions at run-time), whereas hers also includes a fully-fledged regular expression engine.

@@

The safe-regex~\citep{leonhard2021safe} Rust library takes a different approach towards the same goal: it uses Rust macros to statically compile regular expressions and generate arity-specific extractors.
Performance-wise, macros improve over the status quo since they remove the need to compile regular expressions at runtime.
Safety-wise, code generation provides a straightforward solution to safe capturing group extraction.
\end{diff}

### 2022-04-23 thesis.tex
More diff

\noindent
The |Remove| trait takes 3 type parameters: 2 inputs, |V| (the element to remove) and |Ps| (the list), and one output, |Out| (the list, with the element removed).
The two implicit definitions, |casehead| and |casetail|, correspond to the base case and the recursive case of the list removal operation, respectively.
The right-hand side of those definitions is devoid of meaning, instances of the |Remove| trait merely act as placeholders (the type parameters of those constructor calls are inferred from their expected type).
The |Remove| trait is intended to be used as an implicit parameter, with constrained input types, and an unconstrained output type.

>>>>

\noindent
The |Remove| trait takes 3 type parameters: 2 inputs, |V| (the element to remove) and |Ps| (the list), and one output, |Out| (the list, with the element removed).
The two implicit definitions, |casehead| and |casetail|, correspond to the base case and the recursive case of the list removal operation, respectively.
\begin{diff}
The right-hand side of those definitions is devoid of meaning, instances of the |Remove| trait merely act as placeholders (the type parameters of those constructor calls are inferred from their expected type).
\end{diff}
The |Remove| trait is intended to be used as an implicit parameter, with constrained input types, and an unconstrained output type.
